---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Travis-CI Build Status](https://travis-ci.org/jolars/komadown.svg?branch=master)](https://travis-ci.org/jolars/komadown)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/jolars/komadown?branch=master&svg=true)](https://ci.appveyor.com/project/jolars/komadown)
[![CRAN status](http://www.r-pkg.org/badges/version/komadown)](https://cran.r-project.org/package=komadown)
[![codecov](https://codecov.io/gh/jolars/komadown/branch/master/graph/badge.svg)](https://codecov.io/gh/jolars/komadown)

# komadown

komadown provides [R Markdown](http://rmarkdown.rstudio.com/) templates for the
[KOMA-Script classes](https://komascript.de/) (currently only koma-scartcl).
It is intended only for pdf (LaTeX) output.

It wraps around [bookdown](https://github.com/rstudio/bookdown) to enable
cross-referencing of figures, sections. theorems, and equations.

## Installation

You can install komadown from github with

```{r cran-installation, eval = FALSE}
install.packages("komadown")
```

or install the latest development version from github with

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("jolars/komadown")
```

## License

See the [license](LICENSE).
