---
title: "komatufte"
subtitle: "KOMA-script articles inspired by Tufte"
author:
  - name: "Johan Larsson"
    thanks: "mail@larssonjohan.com"
    id: 1
affil:
  - name: "Lund University, Department of Statistics"
    id: 1
date: "`r Sys.Date()`"
output: 
  komadown::komatufte:
    citation_package: biblatex
bibliography: komadown.bib
automark: yes
vignette: >
  %\VignetteIndexEntry{KOMA-script articles inspired by Tufte}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
  extension: pdf
---

```{r, include = FALSE}
knitr::knit_hooks$set(crop = knitr::hook_pdfcrop)

# Fix subfigs for rmarkdown -> latex
if (identical(knitr:::pandoc_to(), 'latex'))
  knitr::knit_hooks$set(plot = knitr::hook_plot_tex)

knitr::opts_chunk$set(
  echo = TRUE,
  dev = "pdf",
  dev.args = list(pointsize = 8),
  crop = TRUE,
  fig.width = 5,
  fig.height = 3
)

knitr::opts_knit$set(width = 50)

options(digits = 3)
```

# Inspired by Edward Tufte

This style has been inspired by the popular document style of Edward
Tufte, which is used throughout his many excellent texts on
data visualization. The distinguishing feature of this style is its
division into a left-aligned main column and a wide margin on the right
into which text, figures, tables, and more can be arranged.

Unlike the
**tufte** package [@xie2018], however, **komadown** takes a bit of 
freedom with this design.^[Instead of italic roman
headings, for instance, we have opted for bold sans-serif such. We have also
decided for a classical approach to references, using a bibliography
at the end of the text.] Unlike the **scrartcl** style that is
also featured in this package, this template takes a much more 
opinionated approach to the design. There are fewer options for
customization^[It is not, however, unlikely that this will change
in the future.]

# Figures

## Default figure environment

**komatufte** features three different figure environments.
The default figure environment has the figure 
positioned in the main column and figure caption in the margin.
\autoref{fig:figure}, taken from Paul Murrels excellent *R Graphics* 
[@murrell2011], showcases the default
behavior of figures in this template.

```{r figure, fig.cap = "Perspective plot of volcano data set.", fig.width = 6, fig.height = 5}
z <- 2 * volcano
x <- 10 * 1:nrow(z)
y <- 10 * 1:ncol(z)

trans <- persp(x, y, z, zlim = c(0, max(z)), 
               theta = 150, phi = 12, lwd = 0.5,
               scale = FALSE, axes = FALSE)

f <- Map(function(i) lines(trans3d(i$x, i$y, 0, trans)),
         contourLines(x, y, z))
```

## Margin figures

There is also the option to put figures, including caption,
altogether into the margin. This environment is invoked by
specifying `fig.env = "marginfigure"` as a chunk option. This
environment is examplified in \autoref{fig:marginfigure}.

```{r marginfigure, fig.cap = "Survival on the Titanic.", fig.env = "marginfigure", fig.width = 2.5, fig.height = 3}
mosaicplot(Survived ~ Sex + Age, data = Titanic, shade = TRUE,
           main = "")
```

## Wide figures

The last figure environment is *widefigure*, enabled by
setting `fig.env = "widefiure".`

```{r widefigure, fig.env = "widefigure", fig.width = 7, fig.height = 2, fig.cap = "Measurements of the annual flow of the river Nile at Aswan (formerly Assuan), 1871–1970, in $10^8$ $m^3$."}
plot(Nile, ylab = "Flow", xlab = "", col = "steelblue4")
```

# Math

Math is typeset using the **newtxmath** package by Michael Sharpe. Note that
the **mathtools** package is also loaded to make available some extra
functionality if needed.

In the following example, we show the objective function for
the elastic net-penalized version of least-squares regression. 
$$
\min_{\beta_0, \beta}
\left\{
  \frac{1}{n} \sum_{i=1}^n \left(y_i -\beta_0 - \beta^\intercal \mathbf{x}_i \right)^2
  + \lambda \left[(1 - \alpha)||\beta||_2^2 + \alpha||\beta||_1 \right]
\right\}.
$$

# Citations and bibliography

**komatufte** is at its best when **natbib** or **biblatex** is
used. If they are, the citations are placed in a two-column full-width
environment. This is specified in the YAML markdown like this:

```{yaml, eval = FALSE}
---
output: 
  komadown::komatufte:
    citation_package: biblatex
---
```

If the default engine, *pandoc-citeproc*, is used instead, the
bibliography will simply be appended after a regular section
titled *References* and will therefore end up in the 
main column.^[There is a pandoc filter, [pandoc-citeproc-preamble](https://github.com/spwhitton/pandoc-citeproc-preamble) that
seems like a perfect solution to this problem, but I have yet to figure out how
to incorporate its Haskell script into this R package.]

# References


