---
title: "scrartcl"
subtitle: "KOMA-script articles with komadown"
author:
  - name: "Johan Larsson"
date: "`r Sys.Date()`"
automark: yes
colorlinks: yes
fontpack-libertine: yes
classoption:
  - DIV=calc
  - headsepline=true
header:
  - pos: ro
    next: komadown
caption:
  - labelfont=bf
  - labelsep=period
  - font=small
setkomafont:
  - element: descriptionlabel
    commands: \normalfont\scshape\bfseries
KOMAoptions:
  - headings=standardclasses
  - DIV=last
bibliography: komadown.bib
output: komadown::scrartcl
vignette: >
  %\VignetteIndexEntry{KOMA-script articles with komadown}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::knit_hooks$set(crop = knitr::hook_pdfcrop)

# Fix subfigs for rmarkdown
if (identical(knitr:::pandoc_to(), 'latex'))
  knitr::knit_hooks$set(plot = knitr::hook_plot_tex)

knitr::opts_chunk$set(
  echo = FALSE,
  fig.width = 2.3,
  fig.height = 2.1,
  dev = "pdf",
  dev.args = list(pointsize = 8),
  crop = TRUE
)

options(digits = 3)
```

# Motivation

The default \LaTeX\ template for [R Markdown](http://rmarkdown.rstudio.com/) and,
by proxy, **pandoc** take liberties with the original design of \LaTeX\ documents.
The title page is compressed, margins are minimized, and first-line indentation 
is usurped by paragraph spacing. This makes sense for markdown documents that 
are intended for several output formats since consistency between those are 
important to uphold. From a typographical standpoint, however, the design leaves
much to desire.

Without delving into the details of this, I think it suffices to say
that the design of \LaTeX\ documents have been wrought with care
and should only be meddled with if one knows what
they are doing.

This, coincidently, is the case with the
[KOMA-Script bundle](https://komascript.de/), which is a 
reimagining of the original \LaTeX\ classes article, book, letter, and book.
In many ways, the KOMA-Script classes are incontrovertible upgrades to the
original classes, adding a significant deal of functionality to boot. 

In addition to this, **komadown** pulls in a small set of other packages
to handle the different needs that KOMA-Script does not cover. All in all,
the package interface introduces

* an interface to `\KOMAoptions`, which makes the majority of KOMA-Script
accessible to the user,
* automatic or manual headers and footers using yaml metadata,
* caption customization via the **caption** package,
* author and affiliation setups via the **authblk** package, and
* font combinations with excellent support for math (mostly using the 
**newtx** package).

In addition to this, **komadown** also leans on the
[bookdown](https://bookdown.org/) package to provide support for
cross-references of tables, figures, theorems, sections, and equations in
R Markdown syntax.

My minor role in this is to provide a \LaTeX\ template for **pandoc** and
interface for **R** to make it easier for R users to draw on the power of
KOMA-Script (and friends).

# Installation

Install **komadown** by running `install.packages("komadown")`.
If you are running RStudio, this will set you up to begin a KOMA-Script
article with **komadown**'s default template (which this document is based on)
simply by going `File -> New File -> R Markdown` and picking the *scrartcl*
template. The code that makes this work is borrowed from
[rticles](https://CRAN.R-project.org/package=rticles).

Besides installing the package, the only other necessary step is to
include
```yaml
---
output: komadown::scrartcl
---
```
somewhere in the metadata. This will eventually call
`pdf_document2()` from **bookdown** and then \LaTeX\ a `.tex` document,
producing `PDF` output.

# Settings

Most of the settings for **KOMA-Script** are called using a metadata block
called `KOMAoptions`, which takes items in the `element=value` form. 
For instance, this document uses this option to switch headings to versions
similar to the standard classes using
```yaml
---
KOMAoptions:
  - headings=standardclasses
---
```
The reader is referred to the **KOMA-script** manual for information on the many
options available. Of special interest may be the `DIV` argument -- which we
will talk about next.

## Type area calculations

One of the strengths of **KOMA-Script** is its facilities for intelligently 
computing type-area layouts. The core \LaTeX\ classes work well out-of-the-box
if the default font options (size, family, line spread) are used. However,
if the user wants to forego Latin/Computer Modern for another font family 
or simply change its size, these settings no longer fullfill their purpose
of keeping the text at an adequate number of words per line.

The core classes solution to this is to use the **geometry** package
and modify the margins manually, but this might take several attempts to get
right and moreover requires that the user knows to get the proportions right.

**KOMA-Script** solves this automatically using an algorithm. This involes
the `DIV=x` setting, which is implemented in **komadown** via
the metadata block `classoption`. The simplest use of `DIV` is to set it to
`default`.
```yaml
---
classoption:
  - DIV=calc
---
```
This uses a predefined table based on the paper size and default font to arrive
at a type area. 

A more advanced use of `DIV` might be to set it to `calc`, in which case
a `DIV` size is calculated based on the other options given KOMA-Script,
such as `BCOR` (binding correction). `calc` also bases its calculations on the
current font settings, but because the calculations are performed at the
time when the document class is loaded and hence
*before* any font packages are included, it does not make sense to set
`DIV` for this reason alone. To get around this, one can use the
`KOMAoptions` block. `KOMAoptions` are called after font packages are loaded,
and hence if the user specifies `DIV=last` (if `DIV=calc` was set in
`classoption`) or `DIV=calc` here, it will successfully adapt the type area.

## Headers and footers

Headers and footers can be specified via the following syntax.
```yaml
---
header:
  - pos: r
    first: "scrartcl"
    next: "komadown"
footers:
  - pos: l
    next: "\today"
---
```
`pos` gives alignment of the header---one of `r`, `l`, and `c` for right, left,
and center respectively. `first` is the text for the header or footer on the
first page and `next` for the latter pages. First and next are optional but
pos is *not*.

Additionally, the user can set
```yaml
---
automark: yes
---
```
which will then create a running header displaying the current section in
the left spot of the header (which is the setting used in this vignette).

## Captions

Captions can be customized using the `caption` metadata block.
```yaml
---
caption:
  - labelfont: bf
  - labelsep: period
  - font: small
---
```
The reader is referred to documentation for the 
[caption package](http://ftp.acc.umu.se/mirror/CTAN/macros/latex/contrib/caption/caption-eng.pdf)
to read about the various settings.

The default settings (the ones above) produce captions with bold face for
the label and a dot separator (Figure \@ref(fig:histogram)).

```{r histogram, fig.cap = "A plot with the default caption settings.", fig.width = 3, fig.height = 2.2}
set.seed(1)
hist(rnorm(10), col = "grey70", xlab = "", main = "")
```

Additionally, we call the **floatrow** package to setup table floats to place
their captions *above* the floats. Settings from this package are currently
not configurable.

# Cross-references

Cross-references are available in markdown syntax via the **bookdown**
package. The basic syntax is `\@ref(label)` where `label` will differ depending
on context:

Figures
:    `label` will be `fig:id` where `id` is the label given the chunk
the figure is produced by. Note that this requires the `fig.cap` argument
to be set or else there won't be a `\figure` environment to attach the label
to.

Sections
:    Add `{#id}` to the end of the section name and you can
reference it with `\@ref(id)`.

Equations
:    Add `(\#eq:id)` inside the math environment and you will be able to
refer to it by `\@ref(eq:id)`

Theorems
:    Add `label="id"` to the theorem chunk and you will be able to refer to
it by `\@ref(prefix:id)` where prefix depends on the type of theorem. See
[this table](https://bookdown.org/yihui/bookdown/markdown-extensions-by-bookdown.html#tab:theorem-envs)
for a description of the various prefixes.

# Font packages

**komadown** includes a few font packs of selected combinations of
serif, sans-serif, and monospace fonts. All are based on the
tremendous \LaTeX\ package [newtx](https://www.ctan.org/pkg/newtx) from Michael
Sharpe. They are accessed with the metadata option `fontpack-x`, `x` being one
of the options in Table \@ref(tab:fontpack).

Table: (\#tab:fontpack) Customized fontpackages

Font pack            Serif              Sans-serif     Monospace
---------------      --------           ------------   ----------
times (default)      Times New Roman    Helvetica      newtxtt
charter              XCharter (Charter) Cabin          Inconsolata
erewhon              Erewhon (utopia)   Cabin          Inconsolata
libertine            Libertine          Biolinum       Inconsolata

Hence, to use the erewhon font pack, you would need to set the following.
```yaml
---
fontpack-erewhon: yes
---
```
If no font package is specified, the default is set to Latin Modern.

# Font settings

All of the font settings can be set or modified with the metadata blocks
`addtokomafont` and `setkomafont`; the former modifies the current options --
the latter resets them. Both use the subitems `element` and `commands`. 
This document, for instance, using the following
settings to modify description lists to use a roman font instead of the 
**KOMA-Script** default of sans-serif.
```
---
setkomafont:
  - element: descriptionlabel
    commands: \normalfont\scshape\bfseries
---
```
As always, please refer to the **KOMA-Script** manual for a thorough 
take on this.

# Bibliography style

The default bibliography style is the *Vancouver Style* [@patrias_2007], which
has been authored by Michael Berkowitz and been included in the package
(licensed under [CC BY-SA 3.0](https://creativecommons.org/licenses/by-sa/3.0/).
It can be replaced with any `.csl` style using the usual **pandoc** interface or
any configuration via ``biblatex`` or ``natbib`.

# Author blocks

The komadown template includes scripting for author blocks as well; in fact,
you *must* use the author blocks. The settings are somewhat involved compared
to the usual author calls. 
```yaml
author:
  - name: "Johan Larsson"
affil:
  - name: "Lund Unviversity"
```
`affil` is, of course, optional. We use the **authblk** package for this. If you
have several authors you use the `id` tag to match authors to their
affiliations.
```yaml
author:
  - name: "Johan Larsson"
    id: 1
  - name: "John Doe"
    id: 1
affil:
  - name: "Lund University"
    id: 1
```
and the authors will link up with the affiliation.

# Design choices

I've attempted to leave most of the design choices at their default values,
putting most of my personal preferences in the default template instead in
order to make them easy to shake off. However, there are a few things
that I have opted to modify. They are

* table captions are always placed on top,
* `number_sections` in the call to `rmarkdown::pdf_document()` is set to `TRUE`
by default, and
* the default bibliography style is set to the *Uniform Requirements* style 
(also known as Vancouver).

# This document

This document uses the following YAML metadata setup.
```yaml
---
title: "scrartcl"
subtitle: "KOMA-script articles with komadown"
author:
  - name: "Johan Larsson"
    affil: "Lund University"
date: "`r Sys.Date()`"
automark: yes
colorlinks: yes
fontpack-libertine: yes
classoption:
  - DIV=calc
  - headsepline=true
header:
  - pos: ro
    next: komadown
caption:
  - labelfont=bf
  - labelsep=period
  - font=small
setkomafont:
  - element: descriptionlabel
    commands: \normalfont\scshape\bfseries
KOMAoptions:
  - headings=standardclasses
  - DIV=last
output: komadown::scrartcl
---
```

# References
