---
title: "scrartcl"
subtitle: "KOMA-script articles with komadown"
author: "Johan Larsson"
date: "`r Sys.Date()`"
automark: yes
colorlinks: yes
header:
  - pos: ro
    first: ""
    next: komadown
caption:
  - format: plain
  - labelfont: bf
  - labelsep: period
  - font: small
output: komadown::scrartcl
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::knit_hooks$set(crop = knitr::hook_pdfcrop)

knitr::opts_chunk$set(
  echo = FALSE,
  fig.width = 2.3,
  fig.height = 2.1,
  dev = "pdf",
  dev.args = list(pointsize = 8),
  crop = TRUE
)

options(digits = 3)
```

# Motivation

The default latex template for [R Markdown](http://rmarkdown.rstudio.com/)
and, by proxy, **pandoc**
take liberties the original design of latex documents. The title
page is compressed, margins are minimized, and first-line indentation
is usurped by paragraph spacing. This makes sense for markdown documents that
are intended for several output formats since consistency between those are
important to uphold. From a typographical standpoint, however, the design
leaves much to wish for. 

I will not delve into the details of this, but I think it suffices to say
that the design of latex documents have been wrought with care
and---at least in my opinion---should only be meddled with if one knows what
they are doing.

This, coincidently, is the case with the
[KOMA-Script bundle](https://komascript.de/), which is a 
reimagining of the original latex classes article, book, letter, and book.
In many ways, the KOMA-Script classes are straight-up upgrades to the original
classes and add a significant deal of functionality to boot.

All of this is (and will) not be a part of **komadown**, but the features that
are include

* automatic type area calculations using the option `DIV` and
* automatic or manual headers and footers using yaml metadata.

In addition to this, **komadown** also gives the user the option to 
customize captions via the [caption](https://www.ctan.org/pkg/caption) package.

**komadown** also includes the possibility for cross-references by locking
in to the [bookdown](https://bookdown.org/) package.

My only part in this is to provide a template and interface for **R** to make
it easier for users to interface with these excellent packages. 

# Installation
Begin by installing **komadown** by running `install.packages("komadown")`.
If you are running RStudio, this will set you up to begin a KOMA-Script
article with **komadown**'s default template (which this document is based on)
simply by going `File -> New File -> R Markdown` and picking the template. The
code that makes this work is borrowed from
[rticles](https://CRAN.R-project.org/package=rticles).

Besides installing the package, the only other necessary step is to
include

```
---
output: komadown::scrartcl
---
```

somewhere in the YAML metadata and you will be good to go.

# Settings

## Headers and footers

Headers and footers can be specified via the following syntax.
```
---
header:
  - pos: r
    first: ""
    next: "KOMA-script articles with komadown"
footers:
  - pos: l
    first: ""
    next: 
---
```
`pos` gives alignment of the header---one of `r`, `l`, and `c` for right, left,
and center respectively. `first` is the text for the header or footer on the
first page and `next` for the latter pages. Note that *none* of these arguments
may be omitted.

Additionally, the user can set
```
---
automark: yes
---
```
which will then create a running header displaying the current section in
the left spot of the header (which is the setting used in this vignette).

## Captions

Captions can be customized using the following settings.
```
---
caption:
  - format: plain
  - labelfont: bf
  - labelsep: period
  - font: small
---
```

These should be self-explanatory; otherwise, the reader is referred to 
the [documentation](http://ftp.acc.umu.se/mirror/CTAN/macros/latex/contrib/caption/caption-eng.pdf)
for the caption package.

The default settings (the ones above) produce captions that look like this:

```{r, fig.cap = "A plot with the default caption settings.", fig.width = 3, fig.height = 2}
hist(rnorm(10))
```


# Design
