---
title: "scrartcl"
subtitle: "KOMA-script articles with komadown"
author: "Johan Larsson"
date: "`r Sys.Date()`"
automark: yes
colorlinks: yes
recalctypearea: yes
header:
  - pos: ro
    first: ""
    next: komadown
caption:
  - format: plain
  - labelfont: bf
  - labelsep: period
  - font: small
bibliography: komadown.bib
output: komadown::scrartcl
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::knit_hooks$set(crop = knitr::hook_pdfcrop)

knitr::opts_chunk$set(
  echo = FALSE,
  fig.width = 2.3,
  fig.height = 2.1,
  dev = "pdf",
  dev.args = list(pointsize = 8),
  crop = TRUE
)

options(digits = 3)
```

# Motivation

The default \LaTeX template for [R Markdown](http://rmarkdown.rstudio.com/) and,
by proxy, **pandoc** take liberties with the original design of latex documents.
The title page is compressed, margins are minimized, and first-line indentation 
is usurped by paragraph spacing. This makes sense for markdown documents that 
are intended for several output formats since consistency between those are 
important to uphold. From a typographical standpoint, however, the design leaves
much to wish for.

I will not delve into the details of this, but I think it suffices to say
that the design of \LaTeX documents have been wrought with care
and---at least in my opinion---should only be meddled with if one knows what
they are doing.

This, coincidently, is the case with the
[KOMA-Script bundle](https://komascript.de/), which is a 
reimagining of the original \LaTeX classes article, book, letter, and book.
In many ways, the KOMA-Script classes are straight-up upgrades to the original
classes and add a significant deal of functionality to boot.

All of this is (and will) not be a part of **komadown**, but the features that
are include

* automatic type area calculations using the option `DIV` and
* automatic or manual headers and footers using yaml metadata.

In addition to this, **komadown** also gives the user the option to 
customize captions via the [caption](https://www.ctan.org/pkg/caption) package
and cross-references in R Markdown by locking
in to the [bookdown](https://bookdown.org/) package.

My only part in this is to provide a template and interface for **R** to make
it easier for users to interface with these excellent packages.

# Installation
Begin by installing **komadown** by running `install.packages("komadown")`.
If you are running RStudio, this will set you up to begin a KOMA-Script
article with **komadown**'s default template (which this document is based on)
simply by going `File -> New File -> R Markdown` and picking the template. The
code that makes this work is borrowed from
[rticles](https://CRAN.R-project.org/package=rticles).

Besides installing the package, the only other necessary step is to
include
```yaml
---
output: komadown::scrartcl
---
```
somewhere in the YAML metadata. This will eventually call
`bookdown::pdf_document2()` and \LaTeX a `.tex` document to produce
`PDF` output.

# Settings

## Type area calculations

One of the strengths of KOMA-Script is its facilities for calculating
type-area sizes intelligently based on the current font settings. 
Larger fonts need more space to reach an adequate number of words per
line whilst smaller fonts need to pull back the margins of the page.
Where users of the core `article` class have to manually figure out
type areas (unless Latin or Computer Modern are used), users of KOMA-Script
have the option to compute this automatically.

KOMA-Script uses an algorithm to decide on the final page size. This involes
the `DIV=x` setting, which is implemented in **komadown** via
YAML metadata. The simplest use of `DIV` is to set it to `default`.
```yaml
---
DIV: default
---
```
This uses a predefined table based on the paper size to arrive at a type area

A more advanced use of `DIV` might be to set it to `calc`, in which case
a `DIV` size is calculated based on the other options given KOMA-Script,
such as `BCOR` (binding correction). `calc` also computes `DIV` based on the
current font settings, but because the calculations are performed
before any font packages area actually loaded, it does not make sense to set
`DIV` for this reason. A better choice is to give the option
`recalctypearea: yes` in the metadata of the R markdown document, in which
case a macro will be called to compute the `DIV` settings after font packages
have been loaded.

## Headers and footers

Headers and footers can be specified via the following syntax.
```yaml
---
header:
  - pos: r
    first: ""
    next: "KOMA-script articles with komadown"
footers:
  - pos: l
    first: ""
    next: 
---
```
`pos` gives alignment of the header---one of `r`, `l`, and `c` for right, left,
and center respectively. `first` is the text for the header or footer on the
first page and `next` for the latter pages. Note that *none* of these arguments
may be omitted.

Additionally, the user can set
```yaml
---
automark: yes
---
```
which will then create a running header displaying the current section in
the left spot of the header (which is the setting used in this vignette).

## Captions

Captions can be customized using the following settings.
```yaml
---
caption:
  - format: plain
  - labelfont: bf
  - labelsep: period
  - font: small
---
```
These should be self-explanatory; otherwise, the reader is referred to 
the [documentation](http://ftp.acc.umu.se/mirror/CTAN/macros/latex/contrib/caption/caption-eng.pdf)
for the caption package.

The default settings (the ones above) produce captions with bold face for
the label and a dot separator (Figure \@ref(fig:histogram)).

```{r histogram, fig.cap = "A plot with the default caption settings.", fig.width = 3, fig.height = 2.2}
set.seed(1)
hist(rnorm(10), col = "grey70", xlab = "", main = "")
```

# Cross-references

Cross-references are made available in markdown syntax via **bookdown**. The
basic syntax is `\@ref(label)`. `label` will differ depending on context.

Figures
:    `label` will be `fig:id` where `id` is the label given the chunk
the figure is produced by. Note that this requires the `fig.cap` argument
to be set or else there won't be a `\figure` environment to attach the label
to.

Sections
:    Add `{#id}` to the end of the section name and you can
reference it with `\@ref(id)`.

Equations
:    Add `(\#eq:id)` inside the math environment and you will be able to
refer to it by `\@ref(eq:id)`

Theorems
:    Add `label="id"` to the theorem chunk and you will be able to refer to
it by `\@ref(prefix:id)` where prefix depends on the type of theorem. See
[this table](https://bookdown.org/yihui/bookdown/markdown-extensions-by-bookdown.html#tab:theorem-envs)
for a description of the various prefixes.

# Font packages

**komadown** includes a couple of collections of fonts. All are based on the
tremendous \LaTeX package [newtx](https://www.ctan.org/pkg/newtx) from Michael
Sharpe. They are accessed with the metadata option fontpack-x, x being one of
the options below

fontpack             serif              sans-serif     monospace
---------------      --------           ------------   ----------
times (default)      Times New Roman    Helvetica      newtxtt
erewhon              Erewhon (utopia)   Cabin          Inconsolata
charter              XCharter (Charter) Cabin          Inconsolata
libertine            Libertine          Biolinum       Inconsolata


Hence, to use the erewhon font pack, you would need to set the following.
```yaml
---
fontpack-erewhon: yes
---
```

# Bibliography style

The default bibliography style is the vancouver [@patrias_2007] style though
any style may be used is a custom .csl style is supplied.

# This document

This document uses the following YAML metadata setup.
```yaml
---
title: "scrartcl"
subtitle: "KOMA-script articles with komadown"
author: "Johan Larsson"
date: "`r Sys.Date()`"
automark: yes
colorlinks: yes
recalctypearea: yes
header:
  - pos: ro
    first: ""
    next: komadown
caption:
  - format: plain
  - labelfont: bf
  - labelsep: period
  - font: small
bibliography: komadown.bib
output: komadown::scrartcl
---
```

# References
